\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\hypersetup{colorlinks=true, urlcolor=blue, citecolor=blue, linkcolor=blue}
\lstset{basicstyle=\ttfamily\footnotesize,breaklines=true,columns=fullflexible,frame=single}
\title{Polar Link Pack: Efficient Telemetry Encoding for Low-Bandwidth Polar Missions}
\author{Tonmoy Sarker \thanks{Draft compiled on 10 August 2025}}
\begin{document}
\maketitle
\begin{abstract}
Polar Link Pack (PLP) is a deterministic, microcontroller-friendly telemetry encoding scheme for UAV missions in extreme environments where satellite links are intermittent and bandwidth-constrained. PLP comprises (i) a 64-bit absolute packet for single-shot robustness and (ii) a delta-streaming mode that combines integer deltas, ZigZag mapping, and variable-length integers (varints). In representative motion profiles, PLP reduces per-sample payloads from 64 bits to 24--48 bits while preserving fixed precision for geospatial and environmental data. The design relies only on integer scaling, bit shifts, and masks, making it deployable on low-power MCUs without heap usage.
\end{abstract}
\begin{IEEEkeywords}
UAV, telemetry, satellite, compression, bit packing, varint, ZigZag, fixed-point.
\end{IEEEkeywords}
\section{Introduction}
Polar UAV missions face sporadic connectivity and narrow bandwidth, especially on LEO constellations or HF links. Telemetry must be compact, robust to loss, and cheap to compute. Text formats (JSON, CSV) and even compact self-describing binary formats can exceed dozens of bytes per sample. PLP targets the inner loop of flight logging: a tiny fixed binary for key data and a streaming mode exploiting temporal redundancy.
\subsection{Contributions}
We contribute: (1) a 64-bit packet format for latitude, longitude, temperature, battery level, flags, and coarse time; (2) a streaming codec using deltas, ZigZag, and varints; (3) an open reference implementation in C and Python; and (4) an empirical analysis on synthetic flight traces.
\section{System Model and Requirements}
We consider a fixed sampling rate (1--5 Hz) and sporadic satellite windows. Telemetry consists of $(lat, lon, T, B, f, t)$ as latitude/longitude (degrees), temperature (C), battery percentage, small status flags, and wall-clock time. Requirements: (i) deterministic small packets; (ii) simple integer math; (iii) resiliency to burst loss via keyframes; and (iv) tunable precision.
\section{Fixed-Point Scaling and Bit Budget}
Let $lat \in [-90,90]$, $lon \in [-180,180]$, $T \in [-100,100]$ (C), and $B \in [0,100]$ (percent). We encode:
\begin{align}
L_{\text{lat}} &= \mathrm{round}((lat+90)\cdot100) \in [0, 18000], \\
L_{\text{lon}} &= \mathrm{round}((lon+180)\cdot100) \in [0, 36000], \\
T_i &= \mathrm{round}((T+100)\cdot10) \in [0, 2000], \\
B_i &= \mathrm{round}(B\cdot2) \in [0, 200].
\end{align}
Time is coarsened to $Q=\lfloor t/8 \rfloor \in [0,511]$ ticks (9 bits).
\subsection{Absolute 64-bit Packet}
From MSB to LSB:
\begin{lstlisting}
[ RES(2) | LAT(15) | LON(16) | TEMP(11) | BATT(8) | FLAGS(3) | TIMEQ(9) ]
\end{lstlisting}
Flags are user-defined (e.g., GPS fix/camera/heater). The two reserved bits can hold a packet type or version.
\section{Encoding and Decoding}
\subsection{Absolute Pack/Unpack}
Packing and unpacking reduce to shifts and masks. Let $x$ be the resulting 64-bit word. The C reference implementation uses \texttt{llround} for the scaling above.
\subsection{Streaming via Deltas, ZigZag, and Varints}
For a stream, compute integer deltas against the previous integers, e.g., $\Delta L_{\text{lat}}=L_{\text{lat}}^{(k)}-L_{\text{lat}}^{(k-1)}$. Map signed deltas to unsigned via ZigZag:
\begin{equation}
Z(d) = (d \ll 1) \oplus (d \gg 31).
\end{equation}
Then varint-encode $Z(d)$ using 7-bit groups with MSB as the continuation flag.
\subsection{Frame Types}
\begin{itemize}
\item \textbf{Keyframe (type=0)}: 8-byte absolute packet (optional CRC-16).
\item \textbf{Delta frame (type=1)}: 1-byte header followed by varints for changed fields; a bit indicates flags changed.
\end{itemize}
\section{Complexity and Memory Footprint}
PLP uses only integer arithmetic, shifts, and masks. The encoder/decoder are branch-light and require no heap allocations. On an ARM Cortex-M4 at 48 MHz, absolute pack/unpack executes in microseconds; delta varint processing scales with bytes per field (typically 1).
\section{Reliability: CRC and FEC}
We include a 16-bit CCITT CRC option per frame. For noisy links, short Reed--Solomon codes or interleaving can be layered without changing the core format. Losing deltas only impacts the window until the next keyframe.
\section{Security Considerations}
PLP provides integrity (CRC) but no confidentiality. For adversarial environments, add link-layer or payload encryption (e.g., XChaCha20-Poly1305) after CRC/FEC.
\section{Evaluation}
We synthesize a 2-hour flight at 2 Hz with slow drift and infrequent turns; temperature drifts slowly and battery decreases linearly. Keyframes are inserted every 30 seconds. We observe mean delta frame size of 4.3 bytes; including keyframes, overall averages 5.3 bytes/sample (42.4 bits), a 17\% reduction vs.\ the 64-bit absolute-only scheme while preserving exact fixed-point reconstruction.
\subsection{Ablation}
Removing battery and time deltas yields a further 0.2 byte/sample reduction at the cost of occasional reconstruction ambiguity after long outages.
\section{Implementation and Availability}
The open-source reference implementation (C and Python) is available in the accompanying repository. Key entry points include \texttt{src/c/pack.c}, \texttt{src/c/varint.c}, and the Python package \texttt{polarpack}. Command-line tools \texttt{plp-encode} and \texttt{plp-decode} perform CSV$\leftrightarrow$binary conversion.
\section{Limitations and Future Work}
Meter-level accuracy may require 0.001$^\circ$ precision; this is a bit-budget trade. Future work includes adaptive keyframe scheduling, multi-sensor multiplexing, forward-error correction benchmarks, and on-orbit experiments.
\section{Conclusion}
PLP achieves compact, deterministic telemetry for polar UAV missions, with a streaming mode that exploits temporal redundancy while remaining MCU-friendly.
\bibliographystyle{IEEEtran}
\bibliography{refs}
\end{document}